Random Generation of Rich \\ Abstract Data Type Values

### 2019-03-12 paper/finalremarks.tex

In the future we aim to extract stronger invariants from the user codebase into
our automatically derived generators, as well as evaluating how introducing
random mutations into structurally complex values might improve testing
performace, specially when generating ``almost correct'' random data.

>>>>

In the future we aim to extract stronger invariants from the user codebase into
our automatically derived generators, as well as evaluating how introducing
random mutations into structurally complex values might improve the testing
performance, specially when generating ``almost correct'' random
data.\looseness=-1

### 2019-03-12 paper/introduction.tex
Added citation to Robby Findler on randomized testing for PLT Redex

% Scenario of random testing
Random testing is a promising approach for finding bugs
\cite{HughesNSA16,HughesPAN16,ArtsHNS15}.
%
\quickcheck \cite{ClaessenH00} is the dominant tool of this sort used by the
Haskell community.
%
It requires developers to specify (i) \emph{testing properties} describing
programs' expected behavior and (ii) \emph{random data generators} based on the
\emph{types} of the expected inputs (e.g., an integer, an string,
etc.). %, e.g., an integer, an string, or an abstract data type (ADT).
%
\quickcheck then generates random test cases and reports violating testing
properties.

@@

%% How to use random generated ADTs
Well-typed ADT values are specially useful when testing programs which expect
highly structured inputs like compilers \cite{Palka11,MidtgaardJKNN17}.
%
Generating ADT values also proves fruitful when looking for vulnerabilities in
combination with fuzzers \cite{GriecoCB16,grieco2017}.
%
%% Problem
%
Despite these success stories, ADT type-definitions do not often capture all the
invariants expected from the data that they are intended to model.
%
As a result, even if random values are well-typed, they might not be built with
enough structure to penetrate into deep software layers.

%% Our proposal
In this work, we identify two different sources of structural information that
can be statically exploited to improve the generation process of ADT values
(Section \ref{sec:sources}).
%
% In this work, we propose a novel improvement in the generation process of ADT
% values by exploiting some static information found in the codebase.
%
Then, we show how to capture this information into our (automatically) derived
random generators.
%
More specifically, we propose a generation process that is capable of
considering how programs branch on input ADTs values as well as how they get
manipulated by abstract interfaces (Section \ref{sec:hrepcont}).
%
Furthermore, we show how to predict the \emph{expected} distribution of the ADT
constructors, values fitting certain branching patterns, and calls to interfaces
that our random generators produce.
%
For that, we extend some recent results on applying \emph{branching
  processes}\cite{gw1875}---a simple stochastic model conceived to study
population growth (Section \ref{sec:synthesis}).
%
%---to predict the distribution of QuickCheck
%generators\cite{DBLP:conf/haskell/MistaRH18}.
%
We implement our ideas in a tool called \dragenp\footnote{\dragenp is available
  at \url{http://github.com/OctopiChalmers/dragen-struct}}, that is capable of
automatically synthesizing \quickcheck generators which produce \emph{rich ADT
  values}, where the distributions of random values can be adjusted at
compile-time to what developers might want. %based on our predictions.
%
Finally, we provide empirical evaluations showing that including static
information from the user codebase improves the code coverage of two external
applications when tested using random values generated following our ideas
(Section \ref{sec:casestudies}).

>>>>

% Scenario of random testing
Random testing is a promising approach for finding bugs
\cite{HughesNSA16,HughesPAN16,ArtsHNS15}.
%
\quickcheck \cite{ClaessenH00} is the dominant tool of this sort used by the
Haskell community.
%
It requires developers to specify (i) \emph{testing properties} describing
programs' expected behavior and (ii) \emph{random data generators} based on the
\emph{types} of the expected inputs (e.g., integers, strings,
etc.). %, e.g., an integer, an string, or an abstract data type (ADT).
%
\quickcheck then generates random test cases and reports violating testing
properties.

@@

%% How to use random generated ADTs
Well-typed ADT values are specially useful when testing programs which expect
highly structured inputs like compilers
\cite{Palka11,MidtgaardJKNN17,klein2009randomized}.
%
Generating ADT values also proves fruitful when looking for vulnerabilities in
combination with fuzzers \cite{GriecoCB16,grieco2017}.
%
%% Problem
%
Despite these success stories, ADT type-definitions do not often capture all the
invariants expected from the data that they are intended to model.
%
As a result, even if random values are well-typed, they might not be built with
enough structure to penetrate into deep software layers.

%% Our proposal
In this work, we identify two different sources of structural information that
can be statically exploited to improve the generation process of ADT values
(Section \ref{sec:sources}).
%
% In this work, we propose a novel improvement in the generation process of ADT
% values by exploiting some static information found in the codebase.
%
Then, we show how to capture this information into our (automatically) derived
random generators.
%
More specifically, we propose a generation process that is capable of
considering how programs branch on input ADTs values as well as how they get
manipulated by abstract interfaces (Section \ref{sec:hrepcont}).
%
Furthermore, we show how to predict the \emph{expected} distribution of the ADT
constructors, values fitting certain branching patterns, and calls to interfaces
that our random generators produce.
%
For that, we extend some recent results on applying \emph{branching
  processes}\cite{gw1875}---a simple stochastic model conceived to study
population growth (Section \ref{sec:synthesis}).
%
%---to predict the distribution of QuickCheck
%generators\cite{DBLP:conf/haskell/MistaRH18}.
%
We implement our ideas as an extension of the already existing derivation tool
called \dragen\cite{DBLP:conf/haskell/MistaRH18}. We call our extension as
\dragenp\footnote{\dragenp is available at
  \url{http://github.com/OctopiChalmers/dragen-struct}} to make it easy the
distinction for the reader.
%
\dragenp is capable of automatically synthesizing \quickcheck generators which
produce \emph{rich ADT values}, where the distributions of random values can be
adjusted at compile-time to what developers might want.
% based on our predictions.
%
Finally, we provide empirical evaluations showing that including static
information from the user codebase improves the code coverage of two external
applications when tested using random values generated following our ideas
(Section \ref{sec:casestudies}).

### 2019-03-12 paper/randomtesting.tex

In order to generate random ADTs values, most approaches require users to provide
a random data generator for each ADT definition.
%
This is a cumbersome and error prone task that usually \emph{follows closely the
  structure of the ADTs}.
%
For instance, consider the following definition of a \quickcheck random
generator for the type |Html|:
%
\begin{code}
genHtml = sized (\size ->
  if size == 0
  then frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString) ]
  else frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString)
  ,  (4,  Tag     <$> genString <*> smaller genHtml)
  ,  (3,  Join    <$> smaller genHtml <*> smaller genHtml) ])
\end{code} %$
%
We use the Haskell syntax |[]| and |(,)| for denoting lists and pairs of
elements, respectively (e.g., |[(1,2),(3,4)]| is a list of pairs of numbers.)
%
The random generator |genHtml| is defined using \quickcheck's function |sized|
to parameterize the generation process up to an external natural number known as
the \emph{generation size}---captured in the code with variable |size|.
%
This parameter is chosen by the user, and it is used to limit the maximum amount
of recursive calls that this random generator can perform and thus ensuring
the termination of the generation process.
%
When called with a positive generation size, this generator can pick to generate
among any |Html| data constructor with an explicitly \emph{generation frequency}
that can be chosen by the user---in this example, 2, 1, 4 and 3 for |Text|,
|Single|, |Tag|, and |Join|, respectively.
%
When it picks to generate a |Text| or a |Single| data constructor, it also
generates a random |String| value using the standard QuickCheck generator
|genString|.% (\quickcheck provides standard random generators for most base types
% like |String|, |Int|, |Bool|, etc.)
\footnote{%
  The operators |<$>| and |<*>| are used in Haskell to combine values obtained
  from calling random generators and they are not particularly relevant for the
  point being made in this work.}
%$
On the other hand, when it picks to generate a |Join| constructor, it also
generates two independent random sub-expression recursively, decreasing the
generation size by a unit on each recursive invocation (|smaller genHtml|).
%
The case of random generation of |Tag| constructors follows analogously.
%
This random process keeps calling itself recursively until the generation size
reaches zero, where the generator is constrained to pick among terminal data
constructors, being |Text| and |Single| the only possible choices in our
particular case.

>>>>

In order to generate random ADTs values, most approaches require users to provide
a random data generator for each ADT definition.
%
This is a cumbersome and error prone task that usually \emph{follows closely the
  structure of the ADTs}.
%
For instance, consider the following definition of a \quickcheck random
generator for the type |Html|:
%
\begin{code}
genHtml = sized (\size ->
  if size == 0
  then frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString) ]
  else frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString)
  ,  (4,  Tag     <$> genString <*> smaller genHtml)
  ,  (3,  Join    <$> smaller genHtml <*> smaller genHtml) ])
\end{code} %$
%
We use the Haskell syntax |[]| and |(,)| for denoting lists and pairs of
elements, respectively (e.g., |[(1,2),(3,4)]| is a list of pairs of numbers.)
%
The random generator |genHtml| is defined using \quickcheck's function |sized|
to parameterize the generation process up to an external natural number known as
the \emph{generation size}---captured in the code with variable |size|.
%
This parameter is chosen by the user, and it is used to limit the maximum amount
of recursive calls that this random generator can perform and thus ensuring
the termination of the generation process.
%
When called with a positive generation size, this generator can pick to generate
among any |Html| data constructor with an explicitly \emph{generation frequency}
that can be chosen by the user---in this example, 2, 1, 4 and 3 for |Text|,
|Single|, |Tag|, and |Join|, respectively.
%
When it picks to generate a |Text| or a |Single| data constructor, it also
generates a random |String| value using the standard \quickcheck generator
|genString|.% (\quickcheck provides standard random generators for most base types
% like |String|, |Int|, |Bool|, etc.)
\footnote{%
  The operators |<$>| and |<*>| are used in Haskell to combine values obtained
  from calling random generators and they are not particularly relevant for the
  point being made in this work.}
%$
On the other hand, when it picks to generate a |Join| constructor, it also
generates two independent random sub-expressions recursively, decreasing the
generation size by a unit on each recursive invocation (|smaller genHtml|).
%
The case of random generation of |Tag| constructors follows analogously.
%
This random process keeps calling itself recursively until the generation size
reaches zero, where the generator is constrained to pick among terminal data
constructors, being |Text| and |Single| the only possible choices in our
particular case.

### 2019-03-12 paper/tikz/hrep.tex
Minor fixes
Added final remarks, tweaks on highlighting

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1.1,xscale=1.1]
%uncomment if require: \path (0,439.4285583496094); %set diagram left start at 0, and has height of 439.4285583496094

@@

% Text Node
\draw (85,133.5) node [scale=0.8]  {$\text{\textbf{module}}\ M$};
% Text Node
\draw (100,159.5) node [scale=0.8]  {$\text{\textbf{data}}\ Html\ =\ ... $};
% Text Node
\draw (107,219.5) node [scale=0.8]  {$div::Html\rightarrow Html$};
% Text Node
\draw (109,233.5) node [scale=0.8]  {$bold::Html\rightarrow Html$};
% Text Node
\draw (86,246.5) node [scale=0.8]  {$hr::Html$};
% Text Node
\draw (275.5,135.5) node [scale=0.8]  {$Html_{Spec}$};
% Text Node
\draw (276,165) node [scale=0.8]  {$Html_{ADT}$};
% Text Node
\draw (276,200) node [scale=0.8]  {$Html_{Patterns}$};
% Text Node
\draw (276,235) node [scale=0.8]  {$Html_{Interface}$};
% Text Node
\draw (276,182.5) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (276,217.5) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (405.5,133.5) node [scale=0.8]  {$Generator$};
% Text Node
\draw (406,178) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (406,217.5) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (514.5,133.5) node [scale=0.8]  {$Html$};
% Text Node
\draw (100,184.5) node [scale=0.8]  {$simpl\ ( ...) \ =\ ... \ $};
% Text Node
\draw (100,198.5) node [scale=0.8]  {$simpl\ ( ...) \ =\ ... \ $};
% Text Node
\draw (202.5,142.5) node [scale=0.7] [align=left] {\textbf{Structure}\\ \textbf{specification}};
% Text Node
\draw (348.5,169.5) node [scale=0.7] [align=left] {\textbf{Generator}\\ \textbf{derivation}};
% Text Node
\draw (466,168.5) node [scale=0.7] [align=left] {\textbf{Random} \\ \textbf{generation}};
% Text Node
\draw (348,245) node [scale=0.7] [align=left] {\textbf{User desired} \\ \textbf{distribution}};

>>>>

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1,xscale=1]
%uncomment if require: \path (0,439.4285583496094); %set diagram left start at 0, and has height of 439.4285583496094

@@

% Text Node
\draw (85,133.5) node [scale=0.8]  {$\text{\textbf{module}}\ \mathrm{M}$};
% Text Node
\draw (100,159.5) node [scale=0.8]  {$\text{\textbf{data}}\ \mathrm{Html}\ =\ ... $};
% Text Node
\draw (107,219.5) node [scale=0.8]  {$\mathrm{div}::\mathrm{Html}\rightarrow \mathrm{Html}$};
% Text Node
\draw (109,233.5) node [scale=0.8]  {$\mathrm{bold}::\mathrm{Html}\rightarrow \mathrm{Html}$};
% Text Node
\draw (84,246.5) node [scale=0.8]  {$\mathrm{hr}::\mathrm{Html}$};
% Text Node
\draw (275.5,140) node [scale=0.8] [align=center] {$\mathrm{Html}$ \text{generation}\\ \text{description}};
% Text Node
\draw (276,165) node [scale=0.8]  {$\mathrm{Html_{ADT}}$};
% Text Node
\draw (276,200) node [scale=0.8]  {$\mathrm{Html_{Patterns}}$};
% Text Node
\draw (276,235) node [scale=0.8]  {$\mathrm{Html_{Interface}}$};
% Text Node
\draw (276,182.5) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (276,217.5) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (405.5,133.5) node [scale=0.8]  {$\text{Generator}$};
% Text Node
\draw (406,178) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (406,217.5) node   {$\boldsymbol{\oplus}$};
% Text Node
\draw (523,140) node [scale=0.8] [align=left] {$\mathrm{Html}$\\\text{test cases}};
% Text Node
\draw (107,184.5) node [scale=0.8]  {$\mathrm{simplify}\ ( ...) \ =\ ... \ $};
% Text Node
\draw (107,198.5) node [scale=0.8] {$\mathrm{simplify}\ ( ...) \ =\ ... \ $};
% Text Node
\draw (202.5,142.5) node [scale=0.7] [align=center] {\textbf{Structure}\\\textbf{specification}};
% Text Node
\draw (348.5,169.5) node [scale=0.7] [align=center] {\textbf{Generator}\\ \textbf{derivation}};
% Text Node
\draw (466,168.5) node [scale=0.7] [align=center] {\textbf{Random} \\ \textbf{generation}};
% Text Node
\draw (348,245) node [scale=0.7] [align=center] {\textbf{User desired} \\ \textbf{distribution}};
